## New York Bight Overarching Model (NYBEM.wrapper) (I)

<span style="color: red;">Dougherty and McKay to write.</span>


### Calculate habitat zones


```{r echo=TRUE}
#Functions for isolating habitat types

##########
#Labels for tidal zones
ZoneTidal <- c("Deep", "Subtidal", "Intertidal", "Upland")

#Create function for isolating tidal zones
set.tidal.zone <- function(bedelev, MLLW, MHHW){
  #Assign zone
  tidalzone <- ifelse(bedelev <= -2, 1,
          ifelse(bedelev <= MLLW, 2,
          ifelse(bedelev <= MHHW, 3, 
          ifelse(bedelev > MHHW, 4, NA))))
  #Send output
  tidalzone
}

##########
#Labels for salinity zones
ZoneSalinity <- c("Marine", "Estuarine", "Fresh")

#Create function for isolating salinity zones
set.salinity.zone <- function(salinity){
  #Assign zone
  salinityzone <- ifelse(salinity >=30, 1, 
                  ifelse(salinity < 30 & salinity >= 0.5, 2, 
                  ifelse(salinity < 0.5 & salinity >= 0, 3, NA)))
  #Send output
  salinityzone
}

##########
#Labels for habitat zones
ZoneHabitat <- c("Upland", "Marine.Deep", "Marine.Subtidal", "Marine.Intertidal", "Estuarine.Subtidal", "Estuarine.Intertidal", "Freshwater")

#Create function for isolating habitat zones
set.habitat.zone <- function(tidalzone, salinityzone){
  #Assign zone
  habitatzone <- ifelse(tidalzone == 4, 1,
          ifelse(salinityzone == 1 & tidalzone == 1, 2,
          ifelse(salinityzone == 1 & tidalzone == 2, 3,
          ifelse(salinityzone == 1 & tidalzone == 3, 4,
          ifelse(salinityzone == 2 & tidalzone == 1 | tidalzone == 2, 5,
          ifelse(salinityzone == 2 & tidalzone == 3, 6,
          ifelse(salinityzone == 3 & tidalzone == 1 | tidalzone == 2 | tidalzone == 3, 7, NA)))))))
  #Send output
  habitatzone
}
```


### Suitability Index Calculator (`SI.calc`)

Insert description here as needed.


### Zone-specific habitat quality functions

Insert description here as needed.


### Rasterize Habitat Patches

__Do we need to reference the spatial coordinate systems? Maybe the R function or package used?__

`patch.rasterizer`: This function rasterizes the output of `patch.mapper` based on a specified grid resolution and aligns that layer with a specified spatial coordinate system. 


```{r echo=TRUE}
#Functions for rasterizing habitat types
patch.rasterizer <- function(patch.mapper.output, grid.resolution, my.coordinate.system){
  #Insert code
}
```


### Other fxns as needed


Numerical code for running the entire model suite


```{r echo=TRUE}
#Function 
NYBEM.wrapper <- function(x, y){
  NYBEM.out <- x + y

  #Send output
  NYBEM.out
}
```
