# Model Verification


Model evaluation is the process for ensuring that numerical tools are scientifically defensible, transparently developed, and appropriate for use in a given application. Chapter 5 presents a broad examination of NYBEM relative to system quality, technical quality, and usability, which aligns with USACE model certification requirements (EC 1105-2-412). This appendix presents additional analyses undertaken to compare model outcomes (habitat zones and quality) with existing habitat for the region, which we refer to as output verification (McKay et al. 2022). This form of verification generally compares model predictions to independent data sets and includes dozens of methods including not only visual and statistical performance but also correcting for pattern recreation and parameterization issues (Bennett et al. 2013). 

The analyses presented in this chapter are an abbreviated version of broader evaluations against empirical data, which are being prepared for peer-reviewed journal articles (Mahan et al. _forthcoming_; Cordero et al. _forthcoming_). These analysis are conducted for the existing condition of the New Jersey Back Bays study system. This section includes three main elements. First, model parameterization is described for this study site. Second, the suite of verification methods are described, which are applied to all six habitat types. Third, model verification results are summarized for the six habitat types.


## New Jersey Back Bays (NJBB) Study System

This section presents a preliminary application of the NYBEM to assess ecosystem condition in the New Jersey Back Bays following three main activities: (1) Collection of hydrodynamic model outcomes, (2) Compilation of other environmental input data sets, and (3) Data post-processes and model execution. Additional description of data and models will be included in the NJBB feasibility study's supplemental EIS. 


### Hydrodynamic Modeling

NYBEM is capable of reading in hydrodynamic conditions from empirical or modeled data sources. However, hydrodynamic models provide the primary basis for USACE planning because of their capacity to span large spatial domains and forecast outcomes over long time horizons with and without project actions. Adaptive Hydraulics (AdH) is the numerical model code applied for hydrodynamic simulations in this study of the New Jersey Back Bays. The two-dimensional (2D) shallow water module of AdH was applied for all simulations, which solves for depth-averaged velocity and salinity throughout the model domain. (More details of the 2D shallow water module of AdH and its computational philosophy and equations are available in Savant et al. 2014 and Savant and Berger 2015.) AdH version 4.6 was applied for this study.

A 2D AdH model was developed and validated for simulation of hydrodynamics and salinity (Details provided in McAlpin and Ross, _draft_). The model was validated to available field data for all parameters. Field data were collected in February 2019 for salinity and discharge/velocity data over a 13-hour tidal cycle at three major inlets – Barnegat, Little Egg, and Great Egg. Field data supported the use of a 2D model as opposed to a three-dimensional (3D) model due to lack of major salinity stratification. The model domain was determined using aerial images and bathymetric/topographic data for the area. The Surface Water Modeling System was used to generate a 2D surface mesh and define material regions for applying specific model features, such as bed roughness. The domain is defined horizontally in Universal Transverse Mercator, zone 18 coordinates with units of meters. Vertically it is based on North American Vertical Datum of 1988 (NAVD88) with units of meters. All data applied to the model are shifted to this datum and coordinate system. Bathymetry data for the model were obtained from several sources: the Coastal Relief Model, sponsor collected hydrographic surveys, and the National Elevation Dataset. These data sets were combined such that the latest data were made a priority as well as data collected at finer resolution. The 2D AdH code can include areas that wet/dry; therefore, elevations were included in the domain up to 2 m NAVD88.

The model domain includes over 9,867 square miles, extending approximately 115 miles along the New Jersey Atlantic Ocean coastline from Lewes, DE, to Manasquan, NJ. The 2D mesh contains 324,881 elements and 165,514 nodes. Resolution varies across the domain as needed to capture changes in terrain and features. For instance, fine resolution may be used in small wetland channels to accurately capture the conveyance of flow in these areas as well as the salinity that migrates upstream in deeper channels. Analyses are computed for a one-year simulation with 2018 boundary conditions. For this simulation, the following variables were output:  

- Bed elevation  
- Water surface elevations: mean lower low water (MLLW), mean tide level (MTL), mean higher high water (MHHW), and 0 to 100% exceedence by 10% (calculated over the 2018 period of record)  
- Salinity levels: mean annual and 0 to 100% exceedence by 10% (calculated over the 2018 period of record)  
- Velocity: mean annual and 0 to 100% exceedence by 10% (calculated over the 2018 period of record)

AdH outputs were interpolated from point data using a spline technique with barriers separating the marine and estuarine data points (i.e., salinity data were not interpolated over the barrier island) using ESRI Spatial Analyst Tools. The interpolated surfaces were then rasterized throughout the model domain. A 10m grid cell size was used as a balance between data resolution from the model outcomes, over- vs. under-parameterization, and application resolution (i.e., the relevant size of wetlands in the region).


### Environmental Data Compilation

Other environmental inputs are required for NYBEM beyond hydrodynamic inputs. Specifically, model inputs are needed for hard bottom substrates, substrate composition, vegetation cover, urban land use composition, shoreline armoring, and vessel traffic. 

The estuarine, subtidal model distinguishes hard bottom ecosystems from soft bottom ecosystems (e.g., oysters vs. SAV/clams, respectively). Hard bottom habitats are generally required for reef environments, and this cultch substrate is often a limiting factor in the region. These substrates tend to be less mobile and persist through time. As such, we use a map of historical oyster beds as a proxy for the presence or absence of hard bottom substrate. Shellfish maps from 1963 were digitized that show recreational, moderate commercial, and high commercial uses, all of which were merged into a single hard bottom data layer.

Substrate composition is a major driving factor in both the estuarine and marine subtidal models. Although many nuanced methods exist for assessing substrate (e.g., grain size, distributional metrics, compaction, etc.), simple composition metrics such as percent sand or fine sediment (silt/clay) suffice for the purpose of NYBEM. Furthermore, other substrate metrics are not available at the scale of the NYBEM model domain. Two data sets were combined for this analysis. First, the [usSEABED](https://www.usgs.gov/programs/cmhrp/science/usseabed) sampling points are available for much of the study area, but underrepresented in the back bay ecosystems. These data were augmented with additional substrate samples from Stockton University and The Nature Conservancy ([TNC](https://maps.coastalresilience.org/newjersey/#)), which focused explicitly on the NJBB area. Sampling points from these two data sets were combined into a single point cloud and composition metrics were interpolated across the region.

Vegetation cover provides an important and rapid measure of the plant community's role in ecological function. Cover classes from the [NLCD](https://www.usgs.gov/centers/eros/science/national-land-cover-database) were used to assess this variable. This data set provides cover as a binary (presence / absence) metric at a 30m cell resolution for the entire Nation. A neighborhood analysis was used to translate this binary metric into a continuous variable. Specifically, a moving window assessment was applied with a 50m buffer distance from each raster cell in the NYBEM domain. Vegetation cover percentage was assessed as the mean cover of the cells within this area. The buffer distance of 50m was chosen as large enough to capture multiple cover measurements relative to the grid resolution but small enough to capture the general role of vegetation in the area.

[NLCD](https://www.usgs.gov/centers/eros/science/national-land-cover-database) data were also used to assess the effects of urbanization on estuarine and marine intertidal systems. All urban land uses with the NLCD (codes 21, 22, 23, and 24) were reclassified as urban, and all other types were classified as non-urban. As was the case with vegetation cover, a moving window assessment was applied with a 100m buffer distance from each raster cell in the NYBEM domain. Neighboring urban land use was assessed as the mean cover of the cells within this area. The buffer distance of 100m was chosen to reflect the pervasive influence of urban processes (e.g., noise, light, and water quality) even at relatively large distances from a given patch.

Shoreline armoring also provides an important metric of human use intensity in the intertidal models. The shoreline armoring data layer is a linear dataset developed by the NOAA Office of Response and Restoration. The layer represents linear areas of the coastal shoreline that are armored (i.e., solid man-made structures, riprap, boulder rubble) as of 2014 (NJ) and 2016 (NY). The Environmental Sensitivity Index (ESI) guidelines define these uses for the preparation, evaluation, and/or response to threats to the coastal environment such as oil spills. The shoreline was originally delineated at mean higher high water (MHHW) using Light Detection and Ranging (LiDAR) and high resolution digital orthophotography datasets.

In subtidal and deepwater ecosystems, vessel density provides an important proxy for other human uses like fishing pressure, navigation, and resource extraction. An Automatic Identification Systems (AIS) vessel traffic density raster dataset was generated by the U.S. Coast Guard Navigation Center, Bureau of Ocean Energy Management, and NOAA Office for Coastal Management from AIS Marine Cadastral data. This layer represents 2019 annual vessel transit counts of all vessels (i.e., commercial and recreational) summarized at a 100 m by 100 m pixel cell resolution. A single vessel transit is counted each time a vessel track passes through, starts, or stops within a 100 m grid cell.  

### Parameter Forecasting

These data sets each represent trade-offs deemed acceptable for this application of the NYBEM. However, different applications could compel use of other data sources. For instance, the NYBEM spatial scale required heavy reliance on regional and national data sets. However, more local applications (e.g., a single bay) could have access to higher resolution data (e.g., locally mapped substrate or shoreline armoring). Data quality and resolution are a common challenge in large scale modeling, and future applications could consider alternative approaches. For this analysis, all data sets were rasterized at the 10m resolution and aligned based on the AdH hydrodynamic data. Table B.1 summarizes all data compilation sources and pre-processing steps. Models were then applied as described in Chapter 4 of this manual.


```{r echo=FALSE}
#Create empty table
TableB.1 <- as.data.frame(matrix(NA, nrow = 0, ncol = 4))
colnames(TableB.1) <- c("Data Layer", "Use in NYBEM", "Source Data", "Post-Processing")

#Specify rows of the table
TableB.1[1,] <- c("Bed elevation (m)", "zone", "AdH", "none")
TableB.1[2,] <- c("Mean lower low water (m)", "zone", "AdH", "none")
TableB.1[3,] <- c("Mean higher high water (m)", "zone", "AdH", "none")
TableB.1[4,] <- c("10th Percentile Salinity (psu)", "zone, est.sub", "AdH", "none")
TableB.1[5,] <- c("Hard Bottom Substrate (presence/absence)", "est.sub", "Historic Shellfish Maps", "none")
TableB.1[6,] <- c("Intertidal slope", "mar.int", "AdH", "Derived from AdH elevation data")
TableB.1[7,] <- c("Relative exposure time", "mar.int", "AdH", "(Hmax-Hmin) / (MHHW-MLLW)")
TableB.1[8,] <- c("Percent of Light Available", "est.sub, mar.sub, mar.deep", "AdH", "exp(-1.39*Hmedian)")
TableB.1[9,] <- c("Change in Velocity (%)", "est.int", "AdH", "Percent change from 2018 baseline")
TableB.1[10,] <- c("Mean annual salinity (psu)", "est.sub", "AdH", "none")
TableB.1[11,] <- c("Duration of salinity greater than 0.5 psu", "fre.tid", "AdH", "Interpolated from observation percentiles")
TableB.1[12,] <- c("Relative depth", "fre.tid, est.int", "AdH", "(MHWW-Hmed) / (Hmax-Hmin)")
TableB.1[13,] <- c("Median velocity (cm/s)", "mar.sub", "AdH", "Unit conversion from AdH")
TableB.1[14,] <- c("Duration of salinity less than 28 psu", "mar.deep", "AdH", "Interpolated from observation percentiles")
TableB.1[15,] <- c("Neighborhood emergent vegetation cover (%)", "fre.tid, est.int, mar.int", "NLCD", "Percent of cells within 50m having a vegetated class")
TableB.1[16,] <- c("Neighborhood urban land use (%)", "est.int, mar.int", "NLCD", "Percent of cells within 100m having urban land classes")
TableB.1[17,] <- c("Distance to shoreline armoring features (m)", "est.in, mar.int", "NOAA ESI", "Compute distance from centroid of cell to nearest feature")
TableB.1[18,] <- c("Substrate sand composition (%)", "est.sub", "usSEABED + TNC", "Point data merging and interpolation")
TableB.1[19,] <- c("Substrate fines composition (%)", "est.sub", "usSEABED + TNC", "Point data merging and interpolation")
TableB.1[20,] <- c("Vessel Density (tracks per year)", "est.sub, mar.sub, mar.deep", "AIS", "2019 tracks intersected with raster")

#Send output table 
rownames(TableB.1) <- NULL
knitr::kable(TableB.1, caption="Table B.1. Summary of data compilation, pre-processing, and forecasting assumptions. Use in NYBEM indicates the module or function as follows: zone mapping (zone), freshwater tidal (fre.tid), estuarine intertidal (est.int), estuarine subtidal (est.sub), marine intertidal (mar.int), marine subtidal (mar.sub), and marine deepwater (mar.deep)", align="c", format = "pipe") 
```


## Methods


NYBEM contains six sub-models addressing different habitat types (Chapter 4). Five sub-models were verified against independent data sets, specifically: freshwater tidal, estuarine intertidal, estuarine subtidal - hard bottom, estuarine subtidal - soft bottom, and marine intertidal. The marine subtidal and marine deepwater models were not verified due to a lack of a single direct empirical analog data set (e.g., a wetland map), although these models will be verified in the future against multiple data sets (e.g., species distribution observations, water quality metrics).

For each sub-model in the analysis, appropriate verification data sets were identified (Table B.2). Key attributes or properties of these data sets were extracted for the analysis. All data were aligned with the NYBEM raster mask described above for direct comparison between NYBEM outputs and verification data.

```{r echo=FALSE}
#Create empty table
TableB.2 <- as.data.frame(matrix(NA, nrow = 0, ncol = 3))
colnames(TableB.2) <- c("NYBEM Sub-Model", "Verification Data", "Attribute or Data Field")

#Specify rows of the table
TableB.2[1,] <- c("Freshwater Tidal (fresh.tid)", "New Jersey Wetlands", "6120 (tidal freshwater marshes), 6141 (phragmites wetlands)")
TableB.2[2,] <- c("Estuarine Intertidal (est.int)", "New Jersey Wetlands", "5411 (open tidal bays), 5412 (tidal mud flat), 6110 (saline marsh)")
TableB.2[3,] <- c("Estuarine Subtidal - Hard Bottom (est.sub.hard)", "TBD", "TBD")
TableB.2[4,] <- c("Estuarine Subtidal - Soft Bottom (est.sub.soft)", "TBD", "TBD")
TableB.2[5,] <- c("Marine Intertidal (mar.int)", "TNC Data?", "TBD")

#Send output table 
rownames(TableB.2) <- NULL
knitr::kable(TableB.2, caption="Table B.2. Summary of verification data used with each NYBEM sub-model", align="c", format = "pipe") 
```

Three families of analyses were undertaken for each sub-model and verification data set. First, observed habitats were translate into presence and absence outcomes for every grid cell in the NJBB domain (i.e., 1 and 0, respectively). Modeled habitat zones were also translate into presence-absence data for comparison. These data were used to verify the efficacy of NYBEM's classification methods. For instance, was a given patch correctly or incorrectly categorized as a freshwater tidal ecosystem in the NYBEM relative to a mapped wetland layer? Model results were collected using confusion matrices, which summarize outcomes relative to true positive, false positive, false negative, and true negative classifications. These summary metrics are provided for the NJBB domain as a whole as well as broken down by inlet and bay regions. Notably, NYBEM's habitat zonation algorithm would be expected to over-predict the extent of a given habitat type because only salinity and tidal limits are considered without consideration of other suitability factors reducing errors; therefore, a high rate of false positives would be anticipated.

The second family of verification analyses focused on NYBEM's habitat suitability predictions. For every patch in a given NYBEM habitat type (e.g., fresh.tid), the overall habitat suitability index (a continuous 0-1 score) was examined relative to presence-absence outcomes from the verification data set. For instance, what was the habitat suitability for all freshwater wetland patches in the verification data set? Ideally, NYBEM would predict high suitability for cells with existing wetlands and low suitability for cells without existing wetlands. Data are summarized for all cells in the NJBB domain with boxplots and violin diagrams.


The third verification analysis examines the overall predictions of NYBEM at large-scales. The total extent of habitat predicted by NYBEM was summarize in both acreage and habitat units (acreage * habitat suitability) for each of eleven inlet in the NJBB domain. Ideally, NYBEM would predict the same order-of-magnitude of habitat in each inlet. Data are summarized in plots of observed and modeled estimates of acreage and habitat units. Again, the NYBEM's habitat zonation would be expected to over-predict habitat extent, although the predictions should be closer to observations for the habitat units.


## Results


NYBEM model verification was conducted separately for each sub-model. The following sections summarize verification analyses and discuss the efficacy of the NYBEM as indicated by these outcomes.

### Freshwater Tidal Model

Note about the input data sets. Presence-absence data and confusion analysis (Table B.3).

Suitability analysis and boxplot / violin plot (Figure B.1).

Bulk prediction across inlets (Figure B.2).

### Estuarine Intertidal Model

Note about the input data sets. Presence-absence data and confusion analysis (Table B.4).

Suitability analysis and boxplot / violin plot (Figure B.3).

Bulk prediction across inlets (Figure B.4).


### Estuarine Subtidal Model

Note about the input data sets. Presence-absence data and confusion analysis (Table B.5).

Suitability analysis and boxplot / violin plot (Figure B.5).

Bulk prediction across inlets (Figure B.6).


### Marine Intertidal Model

Note about the input data sets. Presence-absence data and confusion analysis (Table B.6).

Suitability analysis and boxplot / violin plot (Figure B.7).

Bulk prediction across inlets (Figure B.8).


## Summary

Insert notes on the utility of the NYBEM relative to verification analysis. Note that these analyses provide a blunt view of verification and future analyses will include more nuanced methods such as spatial auto-correlation.

